---
title: "HDAT9600 Final Team Assignment"
author: "Team 6 (Nick Duce, Ruochen Han, Yaru Ding, Kathleen Irena)"
date: "16/08/2022"
output: html_document
---

```{r setup, include=FALSE}
# leave this code here, but feel free to adjust the options or add some more
# see the knitr documentation for details
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=12)

# install all packages needed in task 1&2
library(survminer)
library(eha)
library(survival)
library(bshazard)
library(tidyverse)
library(readr)
library(pROC)
library(arm)
library(faraway)
library(dplyr)
```

## Instructions

* This file (hdat9600assess4_teamX.Rmd) is the R Markdown document in which you need to complete your HDAT9600 final team assignment.
* You should rename the file by replacing 'X' with your team number.
* This assignment is assessed and will count for 30% of the total course marks.
* The assignment comprises two tasks. The first task will focus on logistic regression, and the second task will focus on survival analysis.
* There is no word limit, nor limit on the length of your submitted Rmarkdown (i.e. this file) document.
* The **rendered** html document, were it to be printed, should be **no more** than about 10 A4 pages in length.
* Your html rendered document should include sufficient code chunks, output, and visual display as necessary to support your discursive commentary addressing each task. You may therefore wish to **suppress some of the contents** of your Rmarkdown file from appearing within the rendered document.
* The rendered (html) document will be the submission primarily considered when awarding credit. However, the Rmarkdown file may also be considered as deemed appropriate.
* <strong> Excessively long reports will be penalised </strong> by cessation of awarding of credit after roughly 10 pages of content.
* All tasks are intended to be completed as a team. Have fun discussing the tasks and working together.

Don't hesitate to ask the course convenor for help via OpenLearning. The course instructor are happy to point you in the right direction and to make suggestions, but they won't, of course, complete your assessment for you!


## Data for this assessment
The data used for this assessment consist of records from Intensive Care Unit (ICU) hospital stays in the USA. All patients were adults who were admitted for a wide variety of reasons. ICU stays of less than 48 hours have been excluded. 
The source data for the assessment are data made freely available for the 2012 MIT PhysioNet/Computing for Cardiology Challenge. Details are provided [here](https://physionet.org/challenge/2012/). Training Set A data have been used. The original data has been modified and assembled to suit the purpose of this assessment. While not required for the purposes of this assignment, full details of the preparatory work can be found in the 'hdat9600assess4_final_data_prep' file.
The dataframe consists of 120 variables, which are defined as follows:

#### Patient Descriptor Variables
<li> <em>RecordID:</em>            a unique integer for each ICU stay</li>
<li> <em>Age:</em>                 years</li>
<li> <em>Gender:</em>              male/female</li>
<li> <em>Height:</em>              cm</li>
<li> <em>ICUType:</em>             Coronary Care Unit; Cardiac Surgery Recovery Unit; Medical ICU; Surgical ICU</li>
<li> <em>Length_of_stay:</em>      The number of days between the patient’s admission to the ICU and the end of hospitalisation</li>
<li> <em>Survival:</em>            The number of days between ICU admission and death for patients who died</li>


#### Outcome Variables
<li> <em>in_hospital_death:</em>   0:survivor/1:died in-hospital **this is the outcome variable for Task 1: Logistic Regression**</li>
<li> <em>Status:</em>              True/False **this is the censoring variable for Task 2: Survival Analysis**</li>
<li> <em>Days:</em>                Length of survival (in days) **this is the survival time variable for Task 2: Survival Analysis**</li>


#### Clinical Variables
<em>Use the hyperlinks below to find out more about the clinical meaning of each variable.</em>
The first two clinical variables are summary scores that are used to assess patient condition and risk.

* SAPS-I score - Simplified Acute Physiological Score [Le Gall et al., 1984](http://www.ncbi.nlm.nih.gov/pubmed/6499483)
* SOFA score - Sequential Organ Failure Assessment [Ferreira et al., 2001](http://www.ncbi.nlm.nih.gov/pubmed/11594901)

#### 
The following 36 clinical measures were assessed at multiple timepoints during each patient's ICU stay. For each of the 36 clinical measures, you are given 3 summary variables: a) The minimum value during the first 24 hours in ICU (_min), b) The maximum value during the first 24 hours in ICU (_max), and c) The difference between the mean and the most extreme values during the first 24 hours in ICU (_diff). For example, for the clinical measure Cholesterol, these three variables are labelled 'Cholesterol_min', 'Cholesterol_max', and 'Cholesterol_diff'.

* [Albumin](http://en.wikipedia.org/wiki/Human_serum_albumin)
 (g/dL)
* [ALP](http://en.wikipedia.org/wiki/Alkaline_phosphatase)
 [Alkaline phosphatase (IU/L)]
* [ALT](http://en.wikipedia.org/wiki/Alanine_transaminase)
 [Alanine transaminase (IU/L)]
* [AST](http://en.wikipedia.org/wiki/Aspartate_transaminase)
 [Aspartate transaminase (IU/L)]
* [Bilirubin](http://en.wikipedia.org/wiki/Bilirubin)
 (mg/dL)
* [BUN](http://en.wikipedia.org/wiki/BUN)
 [Blood urea nitrogen (mg/dL)]
* [Cholesterol](http://en.wikipedia.org/wiki/Cholesterol)
 (mg/dL)
* [Creatinine](http://en.wikipedia.org/wiki/Serum_creatinine#Plasma_creatinine)
 [Serum creatinine (mg/dL)]
* [DiasABP](http://en.wikipedia.org/wiki/Diastolic_blood_pressure)
 [Invasive diastolic arterial blood pressure (mmHg)]
* [FiO2](http://en.wikipedia.org/wiki/FIO2)
 [Fractional inspired O<sub>2</sub> (0-1)]
* [GCS](http://en.wikipedia.org/wiki/Glasgow_coma_score)
 [Glasgow Coma Score (3-15)]
* [Glucose](http://en.wikipedia.org/wiki/Serum_glucose)
 [Serum glucose (mg/dL)]
* [HCO3](http://en.wikipedia.org/wiki/Bicarbonate#Diagnostics)
 [Serum bicarbonate (mmol/L)]
* [HCT](http://en.wikipedia.org/wiki/Hematocrit)
 [Hematocrit (%)]
* [HR](http://en.wikipedia.org/wiki/Heart_rate)
 [Heart rate (bpm)]
* [K](http://en.wikipedia.org/wiki/Hypokalemia)
 [Serum potassium (mEq/L)]
* [Lactate](http://en.wikipedia.org/wiki/Lactic_acid)
 (mmol/L)
* [Mg](http://en.wikipedia.org/wiki/Magnesium#Biological_role)
 [Serum magnesium (mmol/L)]
* [MAP](http://en.wikipedia.org/wiki/Mean_arterial_pressure)
 [Invasive mean arterial blood pressure (mmHg)]
* [MechVent](http://en.wikipedia.org/wiki/Mechanical_ventilation)
 [Mechanical ventilation respiration (0:false, or 1:true)]
* [Na](http://en.wikipedia.org/wiki/Serum_sodium)
 [Serum sodium (mEq/L)]
* [NIDiasABP](http://en.wikipedia.org/wiki/Diastolic_blood_pressure)
 [Non-invasive diastolic arterial blood pressure (mmHg)]
* [NIMAP](http://en.wikipedia.org/wiki/Mean_arterial_pressure)
 [Non-invasive mean arterial blood pressure (mmHg)]
* [NISysABP](http://en.wikipedia.org/wiki/Systolic_blood_pressure)
 [Non-invasive systolic arterial blood pressure (mmHg)]
* [PaCO2](http://en.wikipedia.org/wiki/Arterial_blood_gas)
 [partial pressure of arterial CO<sub>2</sub> (mmHg)]
* [PaO2](http://en.wikipedia.org/wiki/Arterial_blood_gas)
 [Partial pressure of arterial O<sub>2</sub> (mmHg)]
* [pH](http://en.wikipedia.org/wiki/Arterial_blood_gas)
 [Arterial pH (0-14)]
* [Platelets](http://en.wikipedia.org/wiki/Platelets)
 (cells/nL)
* [RespRate](http://en.wikipedia.org/wiki/Respiratory_physiology)
 [Respiration rate (bpm)]
* [SaO2](http://en.wikipedia.org/wiki/Arterial_blood_gas)
 [O<sub>2</sub> saturation in hemoglobin (%)]
* [SysABP](http://en.wikipedia.org/wiki/Systolic_blood_pressure)
 [Invasive systolic arterial blood pressure (mmHg)]
* [Temp](http://en.wikipedia.org/wiki/Normal_human_body_temperature)
 [Temperature (&deg;C)]
* [TropI](http://en.wikipedia.org/wiki/Troponin)
 [Troponin-I (&mu;g/L)]
* [TropT](http://en.wikipedia.org/wiki/Troponin)
 [Troponin-T (&mu;g/L)]
* [Urine](http://en.wikipedia.org/wiki/Fluid_balance)
 [Urine output (mL)]
* [WBC](http://en.wikipedia.org/wiki/Reference_ranges_for_blood_tests#Hematology)
 [White blood cell count (cells/nL)]
* Weight (kg)




## Accessing the Data
The data frame can be loaded with the following code:
```
icu_patients_df0 <- readRDS("icu_patients_df0.rds")
icu_patients_df1 <- readRDS("icu_patients_df1.rds")
```
**Note:** icu_patients_df1 is an imputed (i.e. missing values are 'derived') version of icu_patients_df0.  This assessment does not concern the methods used for imputation.


```{r, include=FALSE}
# load two raw datasets "icu_patients_df1" and "icu_patients_df0"
icu_patients_df0 <- readRDS("icu_patients_df0.rds")
icu_patients_df1 <- readRDS("icu_patients_df1.rds")

# data preparation for "icu_patients_df1"
head(icu_patients_df1) # check the basic information of "icu_patients_df1"
nrow(icu_patients_df1)
ncol(icu_patients_df1) 
sum(is.na(icu_patients_df1)) 
icu_removed_na <- icu_patients_df1[colSums(is.na(icu_patients_df1))/nrow(icu_patients_df1) < .2]
icu_removed_na <- na.omit(icu_removed_na)
icu_removed_na$SOFA[icu_removed_na$SOFA == -1] <- NA 
icu_removed_na$Length_of_stay[icu_removed_na$Length_of_stay == -1] <- NA 
icu_removed_na <- na.omit(icu_removed_na) # data cleaning 
nrow(icu_removed_na)
ncol(icu_removed_na)
sum(is.na(icu_removed_na)) # check the basic information of "icu_removed_na" 

# data preparation for "icu_patients_df0"
icu_final <- icu_patients_df0[colSums(is.na(icu_patients_df0))/nrow(icu_patients_df0) < .2]
icu_final <- na.omit(icu_final)
icu_final$SOFA[icu_final$SOFA == -1] <- NA 
icu_final$Length_of_stay[icu_final$Length_of_stay == -1] <- NA 
icu_final$Weight_diff[icu_final$Weight_diff == Inf] <- NA
icu_final$Weight_max[icu_final$Weight_max == -Inf] <- NA
icu_final$Weight_min[icu_final$Weight_min == Inf] <- NA
icu_final$Height[icu_final$Height == -Inf] <- NA
icu_final <- na.omit(icu_final)
nrow(icu_final)
ncol(icu_final) 
sum(is.na(icu_final)) # check the basic information of "icu_final" (use for survival analysis)
```

# Task 1
In this task we construct a logistic regression model to find variables in the ICU dataset that predict in-hospital mortality. We performed exploratory data analysis and have included graphs of our descriptive statistics. We then performed significance testing of between surivors and non-survivors to identify variables of interest. Our approach to formulating our model was to initially create a model using all predictors from the dataset and then use step-wise exclusion of non-significant predictors to formulate our final model. We then assessed the model's goodness-of-fit using a receiver-operator characteristic curve and performed model diagnostics to assess whether our model violated any assumptions of generalised linear models. 

## Demographics

```{r, echo=FALSE, fig.show="hold", out.width="50%"}
# demographics of age
ggplot(data = icu_patients_df1) +
  geom_bar(mapping = aes(x = Age)) +
  labs(x="Age",  title="Age of patients in dataset")

# demographics of gender
ggplot(data = icu_patients_df1) +
  geom_bar(mapping = aes(x = Gender)) +
  labs(x="Gender", y="Frequency", title="Gender in the ICU dataset")
```

```{r, include=FALSE}
icu_removed_na$hospital_death <- factor(icu_removed_na$in_hospital_death, labels=c("Survivor","in_hospital_death"))
```

```{r, include=FALSE}
# remove some descriptor and outcome variables
data_frame <- icu_removed_na %>%
       dplyr::select(-RecordID, -Length_of_stay, -Status, -Days, -in_hospital_death)

# remove the hospital_death
icu_removed_na <- icu_removed_na %>%
                 dplyr::select(-hospital_death)
```

## Fitting multivariate logistic regression models

After examining the dataset, we formulate model 1 using all possible predictor variables (output not included for brevity).

```{r, include=FALSE}
# general look of fitting all variables to our model(mod1)
mod1 <- glm(in_hospital_death ~ SAPS1 + SOFA + Age + Albumin_max + Albumin_min + ALP_max + ALP_min + ALT_diff + AST_max + AST_min + Bilirubin_diff + Bilirubin_max + Bilirubin_min + BUN_diff + BUN_max + BUN_min + Cholesterol_diff + Cholesterol_max + Cholesterol_min + Creatinine_diff + Creatinine_max + Creatinine_min + FiO2_min + GCS_max + GCS_min + Glucose_diff + Glucose_max + HCO3_diff + HCO3_max + HCO3_min + HCT_max + HCT_min + HR_diff + HR_max + ICUType + K_diff + Lactate_diff + Lactate_max + Lactate_min + Na_diff + PaCO2_diff + PaCO2_max + PaCO2_min + PaO2_diff + PaO2_max + PaO2_min + pH_diff + Platelets_diff + Platelets_max + Platelets_min + RespRate_diff + RespRate_max + RespRate_min + Temp_diff + Temp_min + TroponinT_max + Urine_diff + Urine_max + Urine_min + WBC_diff + Weight_max + Weight_min, family = binomial(link = 'logit'), data=icu_removed_na)
summary(mod1)
```

All significant predictors are included in model 2. ICU type is non-significant in model 2 so is excluded from model 3.

```{r}
# sig in Mod1 goes to mod2
mod2 <- glm(in_hospital_death ~ SAPS1 + Age + AST_max + Bilirubin_min + BUN_min + Cholesterol_max + Cholesterol_min + GCS_max + ICUType + pH_diff + Platelets_diff + Platelets_max, family = binomial(link = 'logit'), data=icu_removed_na)
summary(mod2)
```

```{r, include=FALSE}
mod3 <- glm(in_hospital_death ~ SAPS1 + Age + AST_max + Bilirubin_min + BUN_min + Cholesterol_max + Cholesterol_min + GCS_max + Platelets_diff + Platelets_max, family = binomial(link = 'logit'), data=icu_removed_na)
summary(mod3)
```

We then compare mod2 to mod3 using an analysis of deviance. mod2 has the lowest AIC at 1224.3 so we choose mod2 as our final model. 

```{r, echo=FALSE}
anova(mod2, mod3,test='Chisq')
```

## Goodness-of-fit and model diagnostics
```{r}
# testing goodness of fit with a ROC curve
# create predicted values
icu_removed_na %>% mutate(predprob=predict(mod2, type="response")) -> icu_pred_mod2

# rOC curve and AUC of precicted values for full model
test_roc_mod2 = roc(icu_pred_mod2$in_hospital_death ~ icu_pred_mod2$predprob, plot = TRUE, print.auc = TRUE)

# model diagnostics
binnedplot(predict(mod2), residuals(mod2))
```

We found an AUC of 0.819, which shows that our model has fairly good goodness of fit our dataset.We do find that we have quite a few values falling outside the 95% confidence interval bounds on our binned residual plot. This tells us that the residuals may not be normally distributed (and appear to have a negative proponderence) and as such violate the assumption of normality for residuals.

```{r}
icu_patients_df0 <- readRDS("icu_patients_df0.rds")
icu_removed_na2 <- icu_patients_df0[colSums(is.na(icu_patients_df0))/nrow(icu_patients_df0) < .2]
icu_removed_na2 <- na.omit(icu_removed_na2)

icu_removed_na2$SOFA[icu_removed_na2$SOFA == -1] <- NA 
icu_removed_na2$Length_of_stay[icu_removed_na2$Length_of_stay == -1] <- NA 
icu_removed_na2 <- na.omit(icu_removed_na2)

mod4 <- glm(in_hospital_death ~ SAPS1 + Age + BUN_min  + GCS_max + ICUType + Platelets_diff + Platelets_max, family = binomial(link = 'logit'), data=icu_removed_na2)
summary(mod4)
```
When we fit our model to the unimputed dataset, we find that we cannot use some variables as they have large proportions of missing data. This leaves our mdoel with SAPS1, Age, BUN_min, ICUType and platelets_diff and maximum as predictors.


### Task 1 summary
In this task we created a model using the most signficant predictors of in hospital we could find in the ICU dataset. Our final model identified SAPS1, Age, AST_max, Bilirubin_min, BUN_min, Cholesterol_max, Cholesterol_min, GCS_max, Platelets_diff, Platelets_max as significant predictors of in hospital mortality. We found that our model had good goodness of fit to the data with an AUC of 0.819 however, there is a concern that our model may violate the assumption of normality in residuals when examining our binned residual plot.




# Task 2

In this task, we performed exploratory data analysis of sex  and ICU type differences using non-parametric Kaplan-Meier estimations of survival. We then created univariate Cox proportional hazards models for all variables in the ICU dataset. From this, we created a multivariate Cox porportional hazards model using all significant univariate predictors, model 1. We again selected out significant predictors from model 1 to create model 2. Model 3 is a multivariate Cox PH model using all variables as predictors and model 4 uses only significant predictors from model 3. We then compare models 2, 3 and 4 and selected model 2 as our final model as it had the lowest Akaike Information Criterion (AIC). We then tested the proportional hazards assumption of our model and finally refit model 2 to the unimputed dataset. 

## Exploratory data anlysis of gender and ICU type on survival

```{r, include=FALSE}
# choose "gender" variable and comparing survival functions between male and female
om.fitbygender <- survfit( Surv(Days, Status) ~ Gender, data = icu_removed_na)

# Carry out the log-rank tests for differences in survival experience by gender
om.fitbygender2 <- coxph( Surv(Days, Status) ~ Gender, data = icu_removed_na) 
summary(om.fitbygender2)

# choose "ICUType" variable and comparing survival functions between male and female
om.fitbyicutype <- survfit( Surv(Days, Status) ~ ICUType, data = icu_removed_na) 

# carry out the log-rank tests for differences in survival experience by icutype
om.fitbyicutype4 <- coxph( Surv(Days, Status) ~ ICUType, data = icu_removed_na) 
summary(om.fitbyicutype4)
```
```{r, echo=FALSE, fig.show="hold", out.width="50%"}
ggsurvplot(om.fitbygender, pval=TRUE, conf.int=TRUE, title= 'Kaplan-Meier estimate of survival function for Gender') 
ggsurvplot(om.fitbyicutype, pval=TRUE, conf.int=TRUE, title= 'Kaplan-Meier estimate of survival function for ICUType') 
```

From above we can find that the variable 'Gender' is not significant, the variable 'ICUType' is strongly significant just on the level 'ICUTypeCardiac Surgery Recovery Unit' and slightly significant on the level 'ICUTypeSurgical ICU'.

## Formulating univariate models

We formulated univariate models of all predictor variables (output suppressed for brevity). Significant predictors were included in our multivariate model 1. 

```{r, eval=FALSE}
# fitting the univariate Cox proportional hazards models for "icu_removed_na"
covariates <- c("Length_of_stay", "SAPS1", "SOFA", "Age", "Albumin_diff", "Albumin_max", "Albumin_min", "ALP_diff", "ALP_max", "ALP_min", "ALT_diff", 
"ALT_max", "ALT_min", "AST_diff", "AST_max", "AST_min", "Bilirubin_diff", "Bilirubin_max", "Bilirubin_min", "BUN_diff", "BUN_max", "BUN_min", "Cholesterol_diff", "Cholesterol_max", "Cholesterol_min", "Creatinine_diff", "Creatinine_max", "Creatinine_min", "FiO2_diff", "FiO2_max", "FiO2_min", "GCS_diff", "GCS_max", "GCS_min", "Gender", "Glucose_diff", "Glucose_max", "Glucose_min", "HCO3_diff", "HCO3_max", "HCO3_min", "HCT_diff", "HCT_max", "HCT_min", "HR_diff", "HR_max", "HR_min", "ICUType", "K_diff", "K_max", "K_min", "Lactate_diff", "Lactate_max", "Lactate_min", "MAP_diff", "MAP_max", "MAP_min", "Mg_diff", "Mg_max", "Mg_min", "Na_diff", "Na_max", "Na_min", "PaCO2_diff", "PaCO2_max", "PaCO2_min",
"PaO2_diff", "PaO2_max", "PaO2_min", "pH_diff", "pH_max", "pH_min", "Platelets_diff", "Platelets_max", "Platelets_min", "RespRate_diff", "RespRate_max", "RespRate_min", "SaO2_diff", "SaO2_max", "SaO2_min", "Temp_diff", "Temp_max", "Temp_min", "TroponinI_diff", "TroponinI_max", "TroponinI_min", "TroponinT_diff", "TroponinT_max", "TroponinT_min", "Urine_diff", "Urine_max", "Urine_min", "WBC_diff", "WBC_max", "WBC_min", "Weight_diff", "Weight_max","Weight_min")

univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(Days, Status)~', x)))
                        
univ_models <- lapply( univ_formulas, function(x){coxph(x, data = icu_removed_na)})
univ_results <- lapply(univ_models,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coeficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          HR <- paste0(HR, " (", 
                                       HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test", 
                                        "p.value")
                          return(res)
                          return(exp(cbind(coef(x),confint(x))))
                         })
```

## Fitting multivariate models

Model 1 is a multivariate Cox PH model formulated using all significant univariate predictors. We then drop non-significant predictors from model 1 to create model 2. The summary of model 2 is shown below. 

```{r, include=FALSE}
# pick variables with p-value <0.005 from the univariate Cox model as model 1
cox.mv1 <- coxph( Surv(Days, Status) ~ Length_of_stay + SAPS1 + SOFA + Age + Albumin_max + Albumin_min + ALP_max + ALP_min + AST_diff + AST_max + AST_min + Bilirubin_diff + Bilirubin_max + Bilirubin_min + BUN_diff + BUN_max + BUN_min + Cholesterol_max + Creatinine_diff + Creatinine_max + Creatinine_min + FiO2_diff + FiO2_max + GCS_diff + GCS_max + Glucose_diff + Glucose_max + HCO3_diff + HCT_diff + HCT_max + HR_diff + Lactate_diff + Lactate_max + Lactate_min + Na_diff + PaCO2_diff + PaO2_diff + PaO2_max + pH_diff + pH_min + RespRate_max + RespRate_min + Temp_max + Urine_diff + Urine_max + Urine_min + WBC_diff + Weight_max + Weight_min, data = icu_removed_na) 
summary(cox.mv1)
print(drop1(cox.mv1, test="Chisq"))
```

```{r}
cox.mv2 <- coxph(Surv(Days, Status) ~ SAPS1 + Age + Bilirubin_diff + Bilirubin_min + BUN_min + GCS_max + HCT_diff + Lactate_diff + PaO2_max + Weight_max, data = icu_removed_na)
summary(cox.mv2)
```

We then create model 3 which uses all variables as predictors and model 4 uses only the significant predictors from model 3. 

```{r, include=FALSE}
# fit a multivariable Cox model with all variables as model 2
cox.mv3 <- coxph(Surv(Days, Status) ~ Gender + Length_of_stay + SAPS1 + SOFA + Age + ICUType + Urine_diff + Lactate_diff + HR_diff + Bilirubin_diff + Albumin_diff + Albumin_max + Albumin_min + ALP_max + ALP_min + ALT_max + ALT_min + AST_max + AST_min + Bilirubin_max + Bilirubin_min + BUN_max + BUN_min + Cholesterol_max + Cholesterol_min + Creatinine_max + Creatinine_min + FiO2_max + FiO2_min + GCS_max + GCS_min + Glucose_max + Glucose_min + HCO3_max + HCO3_min + HCT_max + HCT_min + HR_max + HR_min + K_max + K_min + Lactate_max + Lactate_min + MAP_max + MAP_min +  Mg_max + Mg_min + Na_max + Na_min + PaCO2_max + PaCO2_min + PaO2_max + PaO2_min + pH_max + pH_min + Platelets_max + Platelets_min + RespRate_max + RespRate_min + SaO2_max + SaO2_min + Temp_max + Temp_min + TroponinI_max + TroponinI_min + TroponinT_max + TroponinT_min + Urine_max + Urine_min + WBC_max + WBC_min + Weight_max + Weight_min + Albumin_diff + ALP_diff + ALT_diff + AST_diff  + Bilirubin_diff + BUN_diff + Cholesterol_diff + Creatinine_diff + FiO2_diff + GCS_diff + Glucose_diff + HCO3_diff + HCT_diff + HR_diff + K_diff + Lactate_diff + MAP_diff + Mg_diff + Na_diff + PaCO2_diff + PaO2_diff+  Platelets_diff + pH_diff + RespRate_diff + SaO2_diff + Temp_diff + TroponinI_diff + TroponinT_diff + Urine_diff + WBC_diff + Weight_diff, data = icu_removed_na)
summary(cox.mv3)
drop1(cox.mv3, test="Chisq")
```

```{r}
# pick variables that are significant based on the previous cox.mv3 likelihood ratio test as model 4
cox.mv4 <- coxph(Surv(Days, Status) ~ Gender + SAPS1 + SOFA + Age + ICUType + Lactate_diff + Bilirubin_diff + GCS_min + HR_min + Mg_max + Platelets_max + Platelets_min + Weight_max + WBC_diff + Weight_diff, data = icu_removed_na)
summary(cox.mv4)
```

At this point we compare model 2 and model 4. The models are not significantly different when examining their analysis of deviance and so we choose model 2 as our final model as it has a lower AIC of 9714.19. 

```{r, echo=FALSE}
print(anova(cox.mv2,cox.mv4))

extractAIC(cox.mv2)
extractAIC(cox.mv4)
```

## Model diagnostics

We tested the proportional hazard assumption for our final model (model 2). 

```{r}
(prop.mvsig <- cox.zph(cox.mv2))
ggcoxzph(prop.mvsig)
```

From these plots, it appears that the curves crossed and the PH assumption may be violated. Recall that the proportional hazard assumption is supported by a non-significant relationship between residuals and time, while a significant relationship favours the alternative of non-constant hazards. The global test is statistically significant (p-value = 9.3e-07) and therefore the proportional hazards assumption is violated. 

## Re-fitting model 2 to the unimputed dataset

Here, we refit model 2 to the unimputed dataset. 

```{r}
cox.mvfinal <- coxph(Surv(Days, Status) ~ SAPS1 + Age + BUN_min + GCS_max + HCT_diff + Weight_max, data = icu_final)
summary(cox.mvfinal)
```
Final model does not include Bilirubin, Lactate, and PaO2 variables due to the remaining available variables after data cleaning of icu_patients_df0. From the output we can see that the final model fits to "icu_final" and only three variables 'Age', 'BUN_min' and 'GCS_max' are significant. These three variables are definitely important in mortality prediction: a study has mentioned BUN can be used as an independent predictor of mortality in ICU patients [[Wernly et al., 2018](https://pubmed.ncbi.nlm.nih.gov/29758935/)], the same finding is also applied for GCS [[Hassan & Liu, 2020](https://journals.lww.com/ccmjournal/Fulltext/2020/01001/747__PROSPECTIVE_EVALUATION_OF_EARLY_GLASGOW_COMA.710.aspx)], and there have been numerous evidence on how patient's survival rate is correlated to to Age

## Task 2 summary
In this task we created a model using the most signficant predictors of in hospital we could find in the ICU dataset. Our final model identified SAPS1, Age, Bilirubin_diff, Bilirubin_min, BUN_min, GCS_max, HCT_diff, Lactate_diff, PaO2_max, Weight_max as significant predictors of in hospital mortality. We found that our model had good goodness of fit to the data, there is a concern that there appears to be a violation of the proportional hazards model. When we fitted the df1 data with the final model, not all variables showed significant effects, with the 'Age', 'BUN_min' and 'GCS_max' variables being the most significant.

